Co-methylation

### 2014-09-30 Rmarkdown/comethylation.Rmd
Further work on Eckhardt et al.

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. I will refer to our best guess of "percentage identical methylation" as  $PIM^{*}$ and discuss some of its properties in the next section.

##### $PIM^{*}$

@@

__UP TO HERE__

__TODO: Write up proof in appendix and link to it__

@@

In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases:

__TODO: Derive results for special cases and illustrate them with a plot__

1. $z_{h, i}$ Bernoulli($p_{h}$) and $z_{h, i'}$ Bernoulli($p_{h'}$) and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
2. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
3. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'} > 0.5$, and $cor(z_{h, i}, z_{h', i'}) \neq c$.
4. $z_{h, i}$ and $z_{h', i'}$ both Bernoulli($p$), $p = p_{h, i} = p_{h', i'}$, and $cor(z_{h, i}, z_{h', i'}) = c$.

__TODO START: Re-write and simplify these special case results__
For case 4 we have:

\begin{align}
\label{eq:pim_iid}
PIM_{1}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
          &= (1 - p)^2 + p^2 \\
\end{align}

By differentiating $PIM_{ind}^{*}$ with respect to $p$ we see that $PIM_{iid} \geq 0.5$.

For case 2 we have:

\begin{align}
\label{eq:pim_inid}
PIM_{2}^{*} &= Pr(z_{h, i} = z_{h', i'}) \\
          &= Pr(Pr(z_{h, i} = 0, z_{h', i'} = 0)) + Pr(Pr(z_{h, i} = 1, z_{h', i'} = 1)) \\
          &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \\
\end{align}

where $p_{h}$ (resp. $p_{h'}$) is the genome-wide average CpG methylation level for the sample from which the $h^{th}$ (resp. $h'^{th}) arose. Unlike $PIM_{iid}$, $PIM_{inid}$ may be less than 0.5. In fact, $0 \leq PIM_{inid} \leq 1$ for $0 \leq p_{h}, p_{h'} \leq 1$.
__TODO END: Re-write and simplify these special case results__

The plot of "percentage identical methylation" in Figure 3a of \cite{Eckhardt:2006gh} asymptotes at 0.4, which is less than the theoretical lowerbound under the assumption of indepdence \eqref{eq:pim_ind}. This is likely due to the fact that the amplicons, and thus the $z_{h, i}$, come from a variety of samples and genomic contexts with different average CpG methylation levels, $p$. Alternatively, it may be that it is not $PIM^{*}$ that is plotted in Figure 3a or that the methylation states of two CpGs separated by a large distance are anti-correlated, resulting in $PIM^{*} < 0.5$ (__TODO: Ref equation__).

* Describe some analyses using autocorrelation of beta-values.
* More recently, within-fragment analyses. Describe concept (move from next section to this section).
* Other definitions/methods
  * Lister's mCG correlations
  * co-methylation across samples
  * "vertical" correlations of \cite{Capra:2014uh}

>>>>

My best guess as to the definition of "percentage identical methylation" is that it is the proportion of pairs of methylation calls at different CpGs from (generally) different amplicons where the methylation states are identical, that is, $Pr(z_{h, i} = z_{h', i'})$. I will refer to my best guess of "percentage identical methylation" as $PIM^{*}$, and now turn to some of its properties.

@@

This result follows by substituting $z_{h, i}$ and $p_{h}$ (resp. $z_{h', i'}$ and $p_{h'}$) for $X_{1}$ and $p_{1}$ (resp. $X_{2}$ and $p_{2}$) into the result derived in __TODO: Cite and link to appendix__.

@@

In general, $PIM^{*} \in [0, 1]$. However, there are a few important special cases that are worth discussing:

1. $PIM^{*}_{inid} = (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} \in [0, 1]$: Independent and non-identically distributed (_inid_), i.e., $z_{h, i} \eqd Bernoulli(p_{h})$ and $z_{h, i'} \eqd Bernoulli(p_{h'})$ and with $cor(z_{h, i}, z_{h', i'}) = c = 0$.
2. $PIM^{*}_{did} = (1 - p)^{2} + p^{2} + 2 c p ( 1 - p)$: Dependent and identically distributed (_did_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c \neq 0$.
3. $PIM^{*}_{iid} = (1 - p)^{2} + p^{2}$: Independent and identically distributed (_iid_), i.e., $z_{h, i}, z_{h', i'} \eqd Bernoulli(p)$ and $cor(z_{h, i}, z_{h', i'}) = c = 0$.

As for $PIM^{*}$, $PIM^{*}_{inid} \in [0, 1]$. In contrast, it is easy to show by differentiation that $PIM^{*}_{iid} \geq 0.5$ and that $PIM^{*}_{did} \geq 0.5 (c + 1)$, which is $> 0.5$ (resp. $< 0.5$) for $c > 0$ (resp. $c < 0$). __TODO: Illustrate these special cases with a plot.__

To return to the plot shown in Figure 3a; the "percentage identical methylation" asymptotes at $40 \%$, or $0.4$, which is less than the theoretical lower-bound of $PIM^{*}_{iid}$. If we believed that all samples had the same average CpG methylation levels, $p_{h}$, then we would conclude that the methylation states of two CpGs separated by a large distance are in fact __anti-correlated__ ($c < 0$) for CpGs separated by approximately $1,000$ bp. However, since we know that there is sample-to-sample variability in the average level of CpG methylation, the more likely explanation for this result is that the $p_{h}$ vary between samples and therefore $PIM^{*} < 0.5$ are possible even supposing $c > 0$.

Of course, the above discussion pre-supposes that what is labelled "percentage identical methylation" in Figure 3a is mathematically equivalent to $PIM^{*}$. I believe this to be true but due to the aforementioned deficiencies of the method description it is not possible to verify this derivation.

Quite apart from what exactly is plotted in Figure 3a, the analysis is based on data from only $2,524$ amplicons, which is nowhere near a genome-wide analysis. Furthermore, methylation levels have been aggregated across multiple samples and so the profile in Figure 3a does not represent any one sample but a pool of samples.

#### Other co-methylation analyses using aggregate measurements

#### Previous within-fragment co-methylation analyses

* More recently, within-fragment analyses. Describe concept (move from next section to this section).

#### Other definitions

* Lister's mCG correlations
* co-methylation across samples
* "vertical" correlations of \cite{Capra:2014uh}
